Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2020-01-10 Manuscript/manuscript.Rmd
edits

We analyzed publically available replicate Hi-C data sets from the GM12878 and K562 cell lines (Supplementary Table 1). Topologically associating domains (TADs) were subsequently called using the ARROWHEAD pipeline [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. <!--After removing TADs that exceeded 2 md in length (biologically plausible maximum TAD size [@dixon2012topological; @sofueva2013cohesin]),--> The results presented are for the GM12878 cell line, with K562 results presented in the Supplementary material. We obtained a total of 8052, 7676, 4670, 2349, and 1031 called TADs across each resolution respectively (Table 1). From the called TADs we obtained a total of 15468, 14253, 8363, 4224, and 1883 unique TAD boundaries (Table 2). The resolution-specific approach we developed to create TAD vs non-TAD boundary regions amenable for binary classification (See Methods) resulted in a total of 535363, 267682, 107073, 53537, and 26768 genomic bins. TAD boundary regions were defined as regions on the linear genome where a genomic bin contained a called TAD boundary ($Y = 1$), while non-TAD boundary regions were defined as genomic bins that did not contain a TAD boundary ($Y = 0$) (Figure 1A).

@@

A total of 78 cell type specific functional genomic annotations were used to build the feature space of downstream random forest models (Supplementary Table 1). The annotations could be broken down into three sets including histone modifications (HM), cis-regulatory elements (CRE), and transcription factor binding sites (TFBS). We constructed three different predictor types for each functional genomic annotation (Figure 1B). Counts of overlaps (**Overlap Counts**, OC) were defined as the total number of genomic annotation regions that overlapped within a given genomic bin. Percents of overlaps (**Overlap Percent**, OP) were defined as the proportion of a bin overlapped by an annotation region. **Distance** was defined as log2 transformed number of base pairs between between the centers of bins and regions defining genomic annotations. A log2-transformation of distance was used to account for the skewness present in the distributions of genomic distances (Supplementary Figure 1). These feature engineering steps were performed to improve TAD boundary predictions by accounting for direct (OC), proportional (OP), and indirect (distance) relationships between TAD boundaries and genomic annotations.

![**Illustrating the resolution-specific data construction and feature engineering used for downstream modeling.** (A) Unique boundaries are denoted by the blue dots. The linear genome was subsequently binned into resolution-specific intervals. The response vector **Y** was defined as a genomic bin that either overlapped with a TAD boundary ($Y = 1$) or did not ($Y = 0$). (B) Three types of associations between bins and genomic annotations (green shapes) used to build the feature space of random forest classification models include _Overlap Counts_ (OC), _Overlap Percent_ (OP), and ($log_{2}$) _Distance_. ](Figures/model_construction_and_predictor_types.png)

@@

To identify the combination of data resolution, predictor type, and re-sampling technique necessary for optimal cell line specific TAD boundary prediction, we created an ensemble framework to systematically compare performances of different model inputs (Figure 2). The framework uses data for a particular cell line *L = {GM12878, K562}*, at a given resolution *R = {5 kb, 10 kb, 25 kb, 50 kb, 100 kb}*, using a particular predictor type *P = {OC, OP, Distance}*, and re-sampling technique *B = {None, RUS, ROS, SMOTE}*. Once the model inputs were established, a random forest classifier was trained on chromosome-specific datasets using 5-fold cross validation. The different combinations of inputs were evaluated by assessing the Matthews Correlation Coefficient (MCC) of each model. Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling $2 \times 5 \times 3 \times 4 \times 21 = 2520$) models. <!--Additional performance metrics are reported in the Supplementary materials.-->

@@

We investigated the effect of class imbalance on model performance using three re-sampling techniques. Expectedly, when using data as-is, that is, with class imbalance present, the models performed poorly. Using any of the resampling techniques improved the performance, with SMOTE-resampled data being the most optimal (Figure 3, Supplementary Figure 2). These results suggest that a combination of both random over- and under-sampling (SMOTE) can improve TAD boundary prediction.

We found that, regardless of resampling technique, random forest classifiers built using a distance-type feature space yielded consistently higher MCCs when compared to the overlap count and overlap percent predictor types (Figure 3), and these results were consistent across resolutions and cell types (Supplementary Figure 2). This indicated that models are better able to discriminate between TAD and non-TAD boundary regions when provided with spatially related associations.

>>>>

We analyzed publically available replicate Hi-C data sets from the GM12878 and K562 cell lines (Supplementary Table 1). Topologically associating domains (TADs) were subsequently called using the ARROWHEAD pipeline [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. <!--After removing TADs that exceeded 2 md in length (biologically plausible maximum TAD size [@dixon2012topological; @sofueva2013cohesin]),--> The results presented are for the GM12878 cell line, with K562 results presented in the Supplementary material. We obtained a total of 8052, 7676, 4670, 2349, and 1031 called TADs across each resolution respectively (Table 1). From the called TADs we obtained a total of 15468, 14253, 8363, 4224, and 1883 unique TAD boundaries (Table 2). The resolution-specific approach we developed to create TAD vs non-TAD boundary regions amenable for binary classification (See Methods) resulted in a total of 535363, 267682, 107073, 53537, and 26768 genomic bins. TAD boundary regions were defined as regions on the linear genome where a genomic bin contained a called TAD boundary ($Y = 1$), while non-TAD boundary regions were defined as genomic bins that did not contain a TAD boundary ($Y = 0$) (Figure 1).

@@

A total of 78 cell type specific functional genomic annotations were used to build the feature space of downstream random forest models (Supplementary Table 1). The annotations could be broken down into three sets including histone modifications (HM), cis-regulatory elements (CRE), and transcription factor binding sites (TFBS). We constructed three different predictor types for each functional genomic annotation (Figure 1). Counts of overlaps (**Overlap Counts**, OC) were defined as the total number of genomic annotation regions that overlapped within a given genomic bin. Percents of overlaps (**Overlap Percent**, OP) were defined as the proportion of a bin overlapped by an annotation region. **Distance** was defined as log2 transformed number of base pairs between between the centers of bins and regions defining genomic annotations. A log2-transformation of distance was used to account for the skewness present in the distributions of genomic distances (Supplementary Figure 1). These feature engineering procedures were performed to improve TAD boundary predictions by accounting for direct (OC), proportional (OP), and indirect (distance) relationships between TAD boundaries and genomic annotations.

![**Illustrating the resolution-specific data construction and feature engineering used for downstream modeling.** Unique boundaries are denoted by the blue dots. The linear genome was subsequently binned into resolution-specific intervals. The response vector **Y** was defined as a genomic bin that either overlapped with a TAD boundary ($Y = 1$) or did not ($Y = 0$). Three types of associations between bins and genomic annotations (green shapes) were then used to build the feature space of random forest classification models include _Overlap Counts_ (OC), _Overlap Percent_ (OP), and ($log_{2}$) _Distance_. ](Figures/model_construction_and_predictor_types.png)

@@

To identify the combination of data resolution, predictor type, and re-sampling technique necessary for optimal cell line specific TAD boundary prediction, we created an ensemble framework to systematically compare performances of different model inputs (Figure 2). The framework uses data for a particular cell line *L = {GM12878, K562}*, at a given resolution *R = {5 kb, 10 kb, 25 kb, 50 kb, 100 kb}*, using a particular predictor type *P = {OC, OP, Distance}*, and re-sampling technique *B = {None, RUS, ROS, SMOTE}*. Once the model inputs were established, a random forest classifier was trained on chromosome-specific datasets using 5-fold cross validation. The different combinations of inputs were evaluated by assessing the Matthews Correlation Coefficient (MCC) of each model. Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling $2 \times 5 \times 3 \times 4 \times 21 (= 2520)$ models. <!--Additional performance metrics are reported in the Supplementary materials.-->

@@

Expectedly, when using data as-is, that is, with class imbalance present, the models performed poorly. Using any of the resampling techniques was found to improve model performance, with RUS narrowly out performing SMOTE re-sampling as the most optimal class balancing technique for all resolutions (Figure 3). These results suggest that simple randomly under-sampling alone can drastically improve TAD boundary prediction. <!--These results suggest that a combination of both random over- and under-sampling (SMOTE) can improve TAD boundary prediction.-->

Additionally, we found that regardless of resampling technique, random forest classifiers built using a distance-type feature space yielded consistently higher MCCs when compared to the overlap count and overlap percent predictor types (Figure 3), and these results were consistent across resolutions and cell types (Supplementary Figure 2). This indicated that models are better able to discriminate between TAD and non-TAD boundary regions when provided with spatially related associations.
